% may be add code snippets here to give a preview of some stuff that had to be done inside Postgres

The main implementation challenge of our provenance system is tracking source information for each tuple while still satisfying the general structure and invariants of the highly optimized Postgres executor. In this section, we describe how provenance information is preserved throughout the execution of an arbitrary SQL query.

\subsection{Executor}

Our system aims to track query level provenance, so it only needs to preserve provenence over the course of a single query plan. We make the assumption that at the base nodes of a query origin information can always be read directly off of tuples during the scan \footnote{This assumption breaks temporary tables, but these can be handled has a special case since there is enough additional information in the \texttt{pg\_prov} table. }. When scanning these base-level tuples, we collect the table id, primary key values, and any other provenance info and pack them into a \texttt{ProvInfo} structure. Our goal is to route \texttt{ProvInfo} through the executor until it reaches an insert statement.

At a high level, the executor consists of a collection of node classes, one for each type of plan node. At initialization, it creates a tree of nodes representing the query plan given by the optimizer. Each node has its own state and memory context and they communicate with each other through an iterator interface. The payload of these iterators is the \texttt{TupleTableSlot} structure, which abstracts over various internal tuple representations. In order to provenance information between nodes, we attach a the structures \texttt{ProvInfo} for the current tuple onto this slot.

\begin{verbatim}
struct TupleTableSlot {
  ...
  List <ProvInfo> provinfos;
}
\end{verbatim}

\begin{figure}
  \centering
    
      \begin{tikzpicture} [node distance=0.2cm, sibling distance=1.5cm, level distance=1.5cm]
        \Tree [ .\node[draw, rectangle](i1){Insert}; [ .\node[draw, rectangle](j1){Join}; [ .\node[draw, rectangle] (f1) {Filter};  
        \node[draw, rectangle](s1){Scan}; ]  \node[draw, rectangle] (s2) {Scan};  ] ]   

        \path [->] (s2) edge[thick, bend right = 50, dotted]  (j1);    
        \path [->] (s1) edge[thick, bend left = 50, dotted] (f1);            
        \path [->] (j1) edge[thick, bend right = 50, dotted] (i1);    
        \path [->] (f1) edge[thick, bend left = 50, dotted] (j1);    
        
      \end{tikzpicture}
    
    \caption {Provenance information is attached to tuples as they are propagated through the execution plan. }
\end{figure}

Since all nodes communicate through this common interface, provenance information is always preserved between nodes. Unfortunately, a node itself may perform arbitary operations on the tuples it receives. We cannot hope to preserve provenence through all nodes, for instance stored procedures, but we can keep maintain origin information for standard SQL queries.   
 
\subsubsection{Scans, Filters, and Projections}

The most basic nodes in a query plan are the ones that receive a single input tuple and optionally output a single tuple, e.g. scans, filters, and projections. Postgres implements this style of node as a projection directly on \texttt{TupleTableSlot}. We simply modify this projection function to preserve the \texttt{ProvInfo} from incoming tuple.

\subsubsection{Joins: Hashing and Sorting}

Joins follow a similar interface as scans, they take in a tuple slot and output projected tuples; however, internally join nodes directly access the inner structure of the tuple. Both hash join nodes and sort merge join nodes first convert their input into a set of \texttt{MinimalTuples} before performing the joins. 

The \texttt{MinimalTuple} structure is are direct representation of the tuple data. It is designed to use up minimal memory in cases where many tuples are in kept main memory. When a join is successfully performed, the resulting fields are projected out directly from the \texttt{MinimalTuple} representation. 

In order to preserve provenance, we need to tag every minimal tuple with it provenance information. Fig~\ref{mintup} gives our new minimal tuple representation. In many cases this provenance information will be negligible, but it could be a source of significant slow down. We explore this furthur in Section~\ref{disc}.

This extra tag is the only modification needed to handle joins. When we convert a tuple from a \texttt{TupleTableSlot} to a \texttt{MinimalTuple}, it is tagged with this information. To internal hashing, sorting, and other operations on minimal tuples this information is opaque.   

\begin{figure}
  \centering
  \label{mintup}
  \begin{tikzpicture}[node distance=2cm]
    \node [abstract, rectangle split, rectangle split parts=3, text width=3cm] {
      Header
      \nodepart{second}Tuple\\ Body
      \nodepart{third}Provenance
    };
  \end{tikzpicture}
  \caption {The modified \texttt{MinimalTuple} tagged with provenance infomation.}
\end{figure}





\subsubsection{Aggregates and Grouping}

Aggregate and grouping nodes present an important special case within the executor . Postgres supports general aggregate operations, and so it contains special structures for maintaining aggregate state that break outside of the tuple structure. To handle aggregates conservatively, we assume that every tuple that is scanned as part of an aggregate function should be a part of the provenance information. To implement this aggregation strategy, we augment the internal state maintained for each aggregate group to also include the provenance. When the aggregation is advanced by one tuple, we add this tuple's provenance information to the state information. 


\subsection{Insertion of Provenance}

The insertion node has the task of finalizing the provenance information for the tuples it receives. Upon receiving a tuple, it first inserts the tuple (so that it has a primary key), and then writes the tuple's \texttt{ProvInfo} to permanent storage. 

Provenance storage is implemented as a Postgres catalog table called \texttt{pg\_prov} shown in Fig~\ref{pgprov}. The schema represents the information as a directed graph between tuples. Additionally, we have a table \texttt{pg\_provcommand} to track the command that lead to each provenance relations. Using the catalog for storage facilitates easy analysis and visualization of provenance information, and it opens up the possibility of a special syntax for querying provenance directly. 

\begin{figure}
  \centering
  \label{pgprov}
\texttt{pg\_prov}\\

\begin{tabular}{|lc|}
  \hline
  Attr & Type \\
  \hline
  origin\_tableid & OID \\
  dest\_tableid & OID \\
  origin\_primary\_key & ARRAY \\
  dest\_primary\_key & ARRAY \\
  command\_id & ID \\
  \hline
\end{tabular}

\texttt{pg\_provcommand}\\

\begin{tabular}{|lc|}
  \hline
  Attr & Type \\
  \hline
  id & ID \\
  sql & TEXT \\
  \hline  
\end{tabular}
\caption{Schemas for the two additional catalog tables that hold provenance information.  }

\end{figure}
